Geometric Filters for Metric Spaces

### 2018-11-27 paper/draft.tex
interim

\begin{algorithmic}
\label{alg:ann_blind}
\Function{ApproximateNearestNeighbor}{$P, q, \eps$}
    \State {Initialize array $L$ with a random permutation of $P$}
    \State {$ c \gets \mbox{the first element of }L$}
    \State {Compute $r_i = \dist(q, p_i)$}
    \While {$|L| > 1$}
        \State \Call{Discard\_1}{$L, c$}
        \ForAll {$p_j \in L $}
            \State {Compute $r_j = \dist(p_j, q)$}
            \Comment {Set the worse candidate $w$ to be either $p_i$ or $p_j$}
            \If {$r_j < r_i$ }
                \State {$ c \gets p_j $}
                \State {$ w \gets p_i $}
            \Else
                \State {$ w \gets p_i $}
            \EndIf
            \State \Call{Discard\_2}{$L, w, c$}
        \EndFor
    \EndWhile
    \State \Return {the unique element left in $L$}
\EndFunction
\end{algorithmic}

@@

% First we consider a simplified version of the problem, where the candidate for being an approximate nearest neighbor
% is given, and we only have to verify this. Without loss of generality, assume that the candidate is $p_1$.

\begin{lemma}
Consider the following random procedure: we start with $n$ numbers $[n] = \{\,1,\dots, n\,\}$ and 
choose $\xi_1$ uniformly at random from $[n]$. If $\xi_1 = n$, then we stop,
otherwise we remove $\xi_1$ and all numbers preceding it and from $\{\,\xi_1, \dots, n\,\}$
we choose $\xi_2$. If $\xi_2 = n$, we stop, if $\xi_2 < n$, we repeat the process removing all numbers not greater than $\xi_2$ and
drawing $\xi_3$ from the remaining numbers uniformly at random. The expected number of steps of this procedure
is $O(\log n)$.
\end{lemma}
\begin{proof}
We can rephrase the procedure as a Markov chain with set of states $[n]$ and transition matrix
\begin{equation}
    P = \begin{pmatrix}
         0      & \frac{1}{n-1} & \frac{1}{n-1} & \dots  & \frac{1}{n-1} & \frac{1}{n-1} \\
         0      &  0            & \frac{1}{n-2} & \dots  & \frac{1}{n-2} & \frac{1}{n-2} \\
         0      &  0            & 0             & \dots  & \frac{1}{n-3} & \frac{1}{n-3} \\
         \vdots &  \vdots       & \vdots        & \ddots & \vdots        & \vdots \\
         0      &  0            & 0             & \dots  & \frac{1}{2} & \frac{1}{2} \\
         0      &  0            & 0             & \dots  & 0           & 1           \\
        \end{pmatrix};
\end{equation}
the chain starts from state $1$.
Note that there is one absorbing state $n$, and all other states are transient, and the matrix is already in the canonical
form, meaning that all transient states precede all absorbing states. 
We denote the square submatrix formed by the first $n-1$
rows and $n-1$ columns by $Q$, it is the submatrix corresponding to all transient states. It is known
(\cite{absorbing_markov_chain}) that the matrix $I-Q$ is invertible, let $N$ be its inverse,
\begin{equation}
N = (I- Q)^{-1} = 
\begin{pmatrix}
         1      & -\frac{1}{n-1} & -\frac{1}{n-1} & \dots  & -\frac{1}{n-1}  \\
         0      &  1             & -\frac{1}{n-2} & \dots  & -\frac{1}{n-2}  \\
         0      &  0             & 1             & \dots  & -\frac{1}{n-3}   \\
         \vdots &  \vdots       & \vdots        & \ddots & \vdots            \\
         0      &  0            & 0             & \dots  & -\frac{1}{2}      \\
         0      &  0            & 0             & \dots  & 0           & 1   \\
        \end{pmatrix}^{-1}
\end{equation}
Let $t_i$ denote the expected time
until absorption, if chain starts from a transient state $i = 1, \dots, n-1$; 
vector of $t_i$'s is given by the formula
\[
\begin{pmatrix}
      t_1 \\
      t_2 \\
      \dots \\
      t_{n-1} \\
      \end{pmatrix}
= N \begin{pmatrix}
      1 \\
      1 \\
      \dots \\
      1 \\
      \end{pmatrix}.
\]
The expected number of steps that we are interested in is $t_1$, so we need to calculate the matrix $N$
and prove that the sum of the elements in its first row is $O(\log n)$.
One can easily check that
\[
N  =  \begin{pmatrix}
      1 & \frac{1}{n-1} & \frac{1}{n-2} & \frac{1}{n-3} & \dots  & \frac{1}{3} & \frac{1}{2} \\
      0 &             1 & \frac{1}{n-2} & \frac{1}{n-3} & \dots  & \frac{1}{3} & \frac{1}{2} \\
      0 &             0 &            1  & \frac{1}{n-3} & \dots  & \frac{1}{3} & \frac{1}{2} \\
 \vdots &        \vdots &  \vdots       & \vdots        & \ddots & \frac{1}{3} & \frac{1}{2} \\
      0 &             0 &            0  &            0  &  \dots &            1 & \frac{1}{2} \\
      0 &             0 &            0  &            0  &  \dots &            0 &           1 \\
      \end{pmatrix},
\]
thus $t_1 = 1 + \frac{1}{n-1} + \frac{1}{n-2} + \dots + \frac{1}{2} = O(\log n)$.
\end{proof}

\paragraph{Experimental Results}
We use the same

>>>>

\begin{algorithmic}
\label{alg:ann_blind}
\Function{ApproximateNearestNeighbor}{$P, q, \eps$}
    \State {Initialize array $L$ with a random permutation of $P$}
    \State {$ c \gets \mbox{the first element of }L$}
    \State {Compute $r_i = \dist(q, p_i)$}
    \State \Call{Discard\_1}{$L, c$}
    \While {$|L| > 1$}
        \ForAll {$p_j \in L $}
            \State {Compute $r_j = \dist(p_j, q)$}
            \Comment {Set the worse candidate $w$ to be either $p_i$ or $p_j$}
            \If {$r_j < r_i$ }
                \State {$ c \gets p_j $}
                \State \Call{Discard\_1}{$L, c$}
                \State {$ w \gets p_i $}
            \Else
                \State {$ w \gets p_i $}
            \EndIf
            \State \Call{Discard\_2}{$L, w, c$}
        \EndFor
    \EndWhile
    \State \Return {the unique element left in $L$}
\EndFunction
\end{algorithmic}

@@

%\begin{lemma}
%Consider the following random procedure: we start with $n$ numbers $[n] = \{\,1,\dots, n\,\}$ and 
%choose $\xi_1$ uniformly at random from $[n]$. If $\xi_1 = n$, then we stop,
%otherwise we remove $\xi_1$ and all numbers preceding it and from $\{\,\xi_1, \dots, n\,\}$
%we choose $\xi_2$. If $\xi_2 = n$, we stop, if $\xi_2 < n$, we repeat the process removing all numbers not greater than $\xi_2$ and
%drawing $\xi_3$ from the remaining numbers uniformly at random. The expected number of steps of this procedure
%is $O(\log n)$.
%\end{lemma}
%\begin{proof}
%We can rephrase the procedure as a Markov chain with set of states $[n]$ and transition matrix
%\begin{equation}
%    P = \begin{pmatrix}
%         0      & \frac{1}{n-1} & \frac{1}{n-1} & \dots  & \frac{1}{n-1} & \frac{1}{n-1} \\
%         0      &  0            & \frac{1}{n-2} & \dots  & \frac{1}{n-2} & \frac{1}{n-2} \\
%         0      &  0            & 0             & \dots  & \frac{1}{n-3} & \frac{1}{n-3} \\
%         \vdots &  \vdots       & \vdots        & \ddots & \vdots        & \vdots \\
%         0      &  0            & 0             & \dots  & \frac{1}{2} & \frac{1}{2} \\
%         0      &  0            & 0             & \dots  & 0           & 1           \\
%        \end{pmatrix};
%\end{equation}
%the chain starts from state $1$.
%Note that there is one absorbing state $n$, and all other states are transient, and the matrix is already in the canonical
%form, meaning that all transient states precede all absorbing states. 
%We denote the square submatrix formed by the first $n-1$
%rows and $n-1$ columns by $Q$, it is the submatrix corresponding to all transient states. It is known
%(\cite{absorbing_markov_chain}) that the matrix $I-Q$ is invertible, let $N$ be its inverse,
%\begin{equation}
%N = (I- Q)^{-1} = 
%\begin{pmatrix}
%         1      & -\frac{1}{n-1} & -\frac{1}{n-1} & \dots  & -\frac{1}{n-1}  \\
%         0      &  1             & -\frac{1}{n-2} & \dots  & -\frac{1}{n-2}  \\
%         0      &  0             & 1             & \dots  & -\frac{1}{n-3}   \\
%         \vdots &  \vdots       & \vdots        & \ddots & \vdots            \\
%         0      &  0            & 0             & \dots  & -\frac{1}{2}      \\
%         0      &  0            & 0             & \dots  & 0           & 1   \\
%        \end{pmatrix}^{-1}
%\end{equation}
%Let $t_i$ denote the expected time
%until absorption, if chain starts from a transient state $i = 1, \dots, n-1$; 
%vector of $t_i$'s is given by the formula
%\[
%\begin{pmatrix}
%      t_1 \\
%      t_2 \\
%      \dots \\
%      t_{n-1} \\
%      \end{pmatrix}
%= N \begin{pmatrix}
%      1 \\
%      1 \\
%      \dots \\
%      1 \\
%      \end{pmatrix}.
%\]
%The expected number of steps that we are interested in is $t_1$, so we need to calculate the matrix $N$
%and prove that the sum of the elements in its first row is $O(\log n)$.
%One can easily check that
%\[
%N  =  \begin{pmatrix}
%      1 & \frac{1}{n-1} & \frac{1}{n-2} & \frac{1}{n-3} & \dots  & \frac{1}{3} & \frac{1}{2} \\
%      0 &             1 & \frac{1}{n-2} & \frac{1}{n-3} & \dots  & \frac{1}{3} & \frac{1}{2} \\
%      0 &             0 &            1  & \frac{1}{n-3} & \dots  & \frac{1}{3} & \frac{1}{2} \\
% \vdots &        \vdots &  \vdots       & \vdots        & \ddots & \frac{1}{3} & \frac{1}{2} \\
%      0 &             0 &            0  &            0  &  \dots &            1 & \frac{1}{2} \\
%      0 &             0 &            0  &            0  &  \dots &            0 &           1 \\
%      \end{pmatrix},
%\]
%thus $t_1 = 1 + \frac{1}{n-1} + \frac{1}{n-2} + \dots + \frac{1}{2} = O(\log n)$.
%\end{proof}

\paragraph{Experimental Results}

In order to experimentally evaluate the performance of our algorithm,
we generate random point sets and random query points, and for each query point
run the algorithm 10 times. The average number of distances to the query point
that were actually computed is the measure that we are interested in.

\begin{figure}[ht]
    \label{fig:ann_const_ratio}
    \includegraphics[width=0.7\textwidth]{log_dependency_constant_ratio.png}
    \caption{Ratio $\log(\mbox{computed distances}) / n$ for ANN algorithm. Data is for uniformly distributed
    points.}
\end{figure}

We used the following methods of generating random points:
\begin{enumerate}
    \item Uniform. Points are sampled uniformly at random from the unit cube in $\R^d$.
    \item Normal. Points are sampled from the standard normal distribution.
\end{enumerate}
Query points were sampled from the uniform distribution on the cube $[-10, 10]^d$
and from the normal distribution centered at the origin with scale 100,
thus we get query points that are "inside" the point set and also "outside".
We sample data in dimensions up to 20 and for $\eps \in \Set{0.001, 0.005, 0.01, 0.05, 0.1}$,
the maximal number of points is $30,000$.
In high dimensions the algorithm does not perform as well for very small epsilon,
but for $\eps = 0.01$ even in dimension 20 just about 10 percent of the distances
are computed for 7,000 points (while for 1,000 points more than 0.3 of the distances must be computed).
We note, however, that the constants in asymptotics are very different for different combinations
of distributions of query points and points in the point set.
